
Read the FCSs
```{r}
library(dplyr)
library(tidyr)
library(reshape2)
library(readxl)

library(flowCore)
library(CATALYST)
library(FlowSOM)

library(ggplot2)
library(ggpubr)

library(ComplexHeatmap)
library(cowplot)

#BiocManager::install("diffcyt")
library(diffcyt)
```
#Load FCS and metadata files
```{r}
prefix <- "CyTOF.analysis"
#FCSdir <- "path to cleaned FCS files as in metadata.xlsx sample table"
printGraph <- F
fixed_seed <- 1234

sampleTable <- read_excel("Metadata.xlsx",sheet="Samples") %>%
    mutate(sample_id=paste(patient_id,Tx_Group,timepoint,sep="."))

sampleTable$Tx_Group <- factor(sampleTable$Tx_Group, levels = c("Placebo", "PolyICLC","Resiquimod"))
sampleTable$timepoint <- factor(sampleTable$timepoint, levels = c("Day0", "Day1","Day29"))

View(sampleTable)
table(sampleTable$Tx_Group,sampleTable$timepoint)

fs <- read.flowSet(files=sampleTable$file_name,path=FCSdir,name.keyword = sampleTable$sample_id,transformation=F)

##this is needed so that prepData will include all markers
for(i in 1:length(fs)){
    keyword(fs[[i]])[["$CYT"]] <- "FACS"
}
```


```{r}
panel <- read_excel("Metadata.xlsx",sheet="MarkerList")
all(panel$antigen %in% colnames(fs))
```

```{r}
sce <- prepData(fs, panel, sampleTable, 
                panel_cols = list(channel = "fcs_colname", antigen = "antigen", class = "marker_class"),
                md_cols=list(file = "file_name", id = "sample_id", factors = c("Tx_Group","timepoint","patient_id")))
```

#Print some QC plots
```{r}
if(printGraph){
    p <- plotExprs(sce, color_by = "Tx_Group")
    p$facet$params$ncol <- 6
    ggsave(p, filename = paste(prefix,"QC.by.Tx_Group.pdf",sep="."),
           width=180,
           height=180,
           units="mm")
    p <- plotExprs(sce, color_by = "timepoint")
    p$facet$params$ncol <- 6
    
    ggsave(p, filename = paste(prefix,"QC.by.timepoint.pdf",sep="."),
           width=180,
           height=180,
           units="mm")
    
    p <- pbMDS(sce, color_by = "Tx_Group", label_by = "sample_id")
    ggsave(p, filename = paste(prefix,"MDSplot.Tx_Group.pdf",sep="."),
           width=180,
           height=180,
           units="mm")
    
    p <- pbMDS(sce, color_by = "timepoint", label_by = "sample_id")
    ggsave(p, filename = paste(prefix,"MDSplot.timepoint.pdf",sep="."),
           width=180,
           height=180,
           units="mm")
}

```

#FlowSOM clustering using only Lineage markers (just for testing)
```{r}
set.seed(fixed_seed)
sce <- cluster(sce, features = "type",
               xdim = 10, ydim = 10, maxK = 20, seed = 1234)

if(printGraph){
    p <- plotExprHeatmap(sce, features = "type", 
                         by = "cluster_id", k = "meta20", 
                         bars = TRUE, perc = TRUE)
    class(p)
    
    pdf(file = paste(prefix,"FlowSOM.ConsensusClust.k20.LineageMarkers.pdf",
                     sep="."), width=7.5, height=5)
    draw(p)
    dev.off()
}

if(saveRDS){
    saveRDS(sce,file=paste(prefix,"SCE.clustSOM100.k20.byCellTypeMarkers.rds",sep="."))
}

#sce <- readRDS(file=paste(prefix,"SCE.clustSOM100.k20.byCellTypeMarkers.rds"))
```
# run t-SNE/UMAP on at most 1000/3000 cells per sample
# this make sure that no samples disproportionately affects the clustering
# also saves run time
```{r}
set.seed(fixed_seed)
sce <- runDR(sce, "TSNE", cells = 1e3, features = "type")
sce <- runDR(sce, "UMAP", cells = 1e3, features = "type")


p1 <- plotDR(sce, "TSNE", color_by = "meta20") + 
    theme(legend.position = "none")
p1$layers[[1]]$aes_params$size <- p1$layers[[1]]$aes_params$size/10

p2 <- plotDR(sce, "UMAP", color_by = "meta20")
p2$layers[[1]]$aes_params$size <- p2$layers[[1]]$aes_params$size/10

lgd <- get_legend(p2)
p2 <- p2 + theme(legend.position = "none")
p <- plot_grid(p1, p2, lgd, nrow = 1, rel_widths = c(5, 5, 2))
ggsave(p,filename = paste(prefix,"TSNE.UMAP.byMeta20.pdf",sep="."),
       width=240,
       height=120,
       units="mm")


if(saveRDS){
    saveRDS(sce,file=paste(prefix,"SCE.clustSOM100.k20.byCellTypeMarkers.tSNE.UMAP.rds",sep="."))
    #sce <- readRDS(file=paste(prefix,"SCE.clustSOM100.k20.byCellTypeMarkers.tSNE.UMAP.rds"))
}
```



#FlowSOM clustering using all markers (we use this for subsequent analysis)
```{r}

sce_all <- cluster(sce, features = panel$fcs_colname,
                   xdim = 10, ydim = 10, maxK = 20, seed = 1234)

p <- plotExprHeatmap(sce_all, features = panel$fcs_colname,
                     by = "cluster_id", k = "meta20",
                     bars = TRUE, perc = TRUE)
class(p)

pdf(file = paste(prefix,"FlowSOM.ConsensusClust.k20.AllMarkers.pdf",sep="."),
    width=10,
    height=5)
draw(p)
dev.off()

if(saveRDS){
    saveRDS(sce_all,file=paste(prefix,"SCE.clustSOM100.k20.byAllMarkers.rds",sep="."))
}
# sce_all <- readRDS(file=paste(prefix,"SCE.clustSOM100.k20.byAllMarkers.rds",sep="."))

```

# run t-SNE/UMAP on at most 1000/3000 cells per sample
```{r}
set.seed(fixed_seed)
sce_all <- runDR(sce_all, "TSNE", cells = 1e3, features = "type")
sce_all <- runDR(sce_all, "UMAP", cells = 1e3, features = "type")

p1 <- plotDR(sce_all, "TSNE", color_by = "meta20") + 
    theme(legend.position = "none")
p1$layers[[1]]$aes_params$size <- p1$layers[[1]]$aes_params$size/10

p2 <- plotDR(sce_all, "UMAP", color_by = "meta20")
p2$layers[[1]]$aes_params$size <- p2$layers[[1]]$aes_params$size/10

lgd <- get_legend(p2)
p2 <- p2 + theme(legend.position = "none")
p <- plot_grid(p1, p2, lgd, nrow = 1, rel_widths = c(5, 5, 2))
ggsave(p,filename = paste(prefix,"TSNE.UMAP.byMeta20.allMarkers.pdf",sep="."),
       width=240,
       height=120,
       units="mm")

if(saveRDS){
    saveRDS(sce_all,file=paste(prefix,"SCE.clustSOM100.k20.byAllMarkers.tSNE.UMAP.rds",sep="."))
    sce_all <- readRDS(file=paste(prefix,"SCE.clustSOM100.k20.byAllMarkers.tSNE.UMAP.rds",sep="."))
}
```

#Merge Clusters based on manual annotation
```{r}
merging_table1_file <- "ClusterMerging_AllMarkers.xlsx"
merging_table1 <- read_excel(merging_table1_file,sheet="Mapping",range="A1:B21")
color_table1 <- read_excel(merging_table1_file,sheet="Color")
View(merging_table1)

merging_table1$new_cluster <- factor(merging_table1$new_cluster, 
                                     levels = color_table1$cluster_name)

colorList <- color_table1$color
names(colorList) <- color_table1$cluster_name
colorList

sce_all <- mergeClusters(sce_all, k = "meta20", 
                         table = merging_table1, id = "merging1",overwrite = T)

```


###draw plots
```{r}
p1 <- plotDR(sce_all, "TSNE", color_by = "merging1", k_pal = colorList) + 
    theme(legend.position = "none")
p1$layers[[1]]$aes_params$size <- p1$layers[[1]]$aes_params$size/10

p2 <- plotDR(sce_all, "UMAP", color_by = "merging1", k_pal = colorList)
p2$layers[[1]]$aes_params$size <- p2$layers[[1]]$aes_params$size/10

lgd <- get_legend(p2)
p2 <- p2 + theme(legend.position = "none")
p <- plot_grid(p1, p2, lgd, nrow = 1, rel_widths = c(5, 5, 2))
ggsave(p,filename = paste(prefix,"TSNE.UMAP.byManualAnnot.allMarkers.pdf",sep="."),
       width=240,
       height=120,
       units="mm")

p <- plotExprHeatmap(sce_all, features = panel$fcs_colname,
                     by = "cluster_id", k = "merging1",bars = TRUE, perc = TRUE)

pdf(file = paste(prefix,"Heatmap.ManualAnnot.AllMarkers.pdf",sep="."),
    width=10,
    height=5)
draw(p)
dev.off()

if(saveRDS){
    saveRDS(sce_all,file=paste(prefix,"SCE.clustSOM100.k20.byAllMarkers.tSNE.UMAP.merged.rds",sep="."))
    #sce_all <- readRDS(file=paste(prefix,"SCE.clustSOM100.k20.byAllMarkers.tSNE.UMAP.merged.rds",sep="."))
}

##The clustering and manual merging up to this point, below are visualizations
##Can move to another script
```

